Phylotyper: {\it In silico} predictor of molecular subtypes from gene sequences \\ \large Supplementary Information

### 2017-05-10 manuscript/response_to_reviewers.txt

Although the GitHub page is well laid out and easy enough to follow, please consider including some sort of example to follow along with. As a newcomer to the application it took me a little while to figure out how to get everything exactly right for an initial run. I think a very small example to follow and run that is located early in the GitHub instructions would really help to just get things going.

*Response:
We agree with the reviewer's suggestion that a working example provided for new users will enhance the adoption and usability of Phylotyper.  We have added an "Example" section to the README.md file (which is also used as the landing page for Phylotyper's GitHub repository: https://github.com/superphy/insilico-subtyping/) along with example data that demonstrates how to 1) predict subtypes for schemes already packaged in Phylotyper and 2) how to add and run user-defined subtype schemes.

There lies a slight issue in the "user generated sub-type schemes" in that the results that you get with Phylotyper are highly dependant on the quality of the sub-type scheme provided. Since this tool will likely be used in outbreak situations and analyses, do you have any plans to address the quality of the sub-type scheme that was used? Do you plan on having a section in the GitHub page that allows users to send new sub-type schemes for validation and posting?

*Response*:
The reviewer is correct that Phylotyper's performance is dependent on the input reference sequences used to build the subtype predictor. To mitigate poor prediction from unsuitable or inaccurate inputs, we have included in the build pipeline for the local user-submitted subtype feature a series of checks to automatically test the performance of any submitted subtype scheme. These tests 1) verify that the phylogeny of the provided sequences places the same subtypes into phylogenetic groups and 2) that the subtype can be accurately predicted with the given input sequences by running a leave-one-out cross-valiation. The phylogenetic test empirically determines the distribution of inter-patristic distance between the sequences in the tree with the same subtypes. It uses the distribution to identify subtree groupings that place different subtypes in the same clade that are expected to be the same. Users are notified if any unexpected subtype assignments are detected in the input. In our experience, this test was effective at identifying a limited number of anomylous subtype assignments, for example, incorrectly annotated sequences in the larger subtype scheme. In the cross-validation test, each input reference sequence is withheld from the training set and used to test the predictor.  Users are notified if the subtype scheme has a F1-score below 0.9 from the cross-validation test. We found that this test identified subtypes schemes that have poor predictive power due to the subtype being weakly phylogenetically correlated with the input sequences or subtypes being underrepresented in the training set.
While these checks identify major breakdowns in the requirements for an accurate subtype predictor, we note there is no check to identify missing data, specifically, missing subtypes or key subtypes gene sequences that define the range of the subtype sequence space. For any current general subtype predictor (i.e. not predictors that replicate PCR in silico, for instance), performance is dependent on the reference input set. In dealing with missing data, we demonstrate that Phylotyper is significantly more robust than a direct sequence similarity-based predictor. It is less likely to generate a type-I error when encountering a subtype not found in the training set. The Phylotyper method estimates the uncertainty associated with a subtype prediction. When test sequences that are phylogenetically distant from known reference sequences or in a region of the phylogenetic tree that is rapidly evolving, the Phylotyper score will reflect the degree of uncertaintity. Most often the sequence for a novel or missing subtype will be relatively phylogenetically distant from the other subtype sequences in the tree and its Phylotyper uncertaintity score will reflect this. Only in cases where the novel/missing subtype is closely phylogenetically related to an existing subtype in the training set, will Phylotyper generate a false positive prediction. This property of Phylotyper makes it more robust when working with incomplete training data.

@@

Reviewer: 3

Comments to the Author
It's good to see a typing tool that is based on phylogenetic analysis. As public health practitioners still need to compare WGS data to old typing results, this tool will be useful to them.  Moreover, by putting typing into the framework of phylogenetic analysis, this tool can provide an interesting learning/visualization tool for epidemiologists and other public health practitioners to bridge the gap between subtypes (categorical data) and a tree (continuous data). The tool appears to be well implemented.
Some Specific Comments that the authors should address:
Major:
1) Please comment on what happens when an incomplete loci is found in the WGS data. Does the software warn the users of an partial alignment due to incomplete gene?

*Response:
Complete or nearly complete loci sequences are needed for correctly placing the loci in the phylogenetic tree. To prevent incomplete loci being used, we set coverage cutoffs of 90% in the BLAST detection step. No alleles with less than this coverage to one of the reference set sequences will pass the detection step. Users are notified when no loci I have added a description of the requirement for complete loci sequences and the checks we have built into Phylotyper to the Discussion.

2) Please clarify how novel subtypes/alleles are identified and reported.

*Response
Phylotyper is significantly more robust than a direct sequence similarity-based predictor in handling novel subtypes or alleles. We do not mean that Phylotyper performs unsupervised construction of new subtype schemes (although the underlying method could be adapted for this task), rather, that Phylotyper has significantly lower false positive rates when performing prediction on an allele not in the reference set or an allele for which there is no example with the same subtype in the reference set (i.e. novel subtype). In the case of a novel subtype, Phylotyper would not assign the allele any subtype.  Compare this result with a sequence similarity approach, which, in most cases assign the novel subtype the closest subtype based on sequence.

To clarify this distinction, we have added and modified text.  We have also performed and included a subsequent analysis comparing the accuracy of the BLAST-based approach to the Phylotyper approach when the entire subtype group is withheld from the training set to simulate performance when encountering a novel subtype. This analysis shows that Phylotyper is a marked improvement over the sequence-similarity approach.

3) As there is a trend moving to core-genome or whole-genome MLST typing schemes, please comment if Phylotyper could be scaled up to handle that. How would horizontal gene transfer or recombination affect the Phylotyper results?

### 2017-05-10 manuscript/phylotyper_revised.tex
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotyping  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
Region phylogeny is predictive of the subtype
Linkage disequalibrium
homologous genes
Correct alignment and construction of phylogeny
Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.

@@

To compare the performance of Phylotyper to a sequence similarity-based approach, we ran a leave-one-out cross validation analysis.
For this assessment, we developed a mock sequence-similarity based tool that assigns putative subtypes using BLAST.
The leave-one-out analysis examined the four subtype schemes available in Phylotyper. The results of the analysis indicated that the precision of the Phylotyper method is consistently higher than a top-BLAST hit approach (Supplementary Table S2).

@@

Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type 1 error when encountering a novel subtype. Our empircal testing confirmed this behavior. 
The rate of false positive classifications when we withheld an entire subtype from the training set and used as it the test set was significantly lower than with a straight-forward

>>>>

Currently, subtype schemes for \emph{Escherichia coli} (\textit{E. coli}) are available in the Phylotyper package, and in addition to Stx include intimin and serotype O- and H-types  (Supplementary Table~1).
However, the Phylotyper software also has the capability to add new subtype schemes. 
Creating a new subtype scheme will save the required reference files, allowing newly added schemes to be easily re-run from Phylotyper.
Performance of the Phylotyper method is dependent on the quality of the subtype scheme.
Region / phylogeny is predictive of the subtype
Linkage disequalibrium
homologous genes
Correct alignment and construction of phylogeny
Checks are built-in to the new subtype pipeline to ensure the assumptions of the Phylotyper approach are not violated.

@@

Phylotyper marks a progression from the sequence-similarity approach widely used in \it{in silico} subtyping strategies.
It's performance extends beyond allele matching, to perform accurate inference of subtypes for input gene sequences that not in the training data or reference database.
To compare Phylotyper to a sequence similarity-based approach, we ran two validations that looked at how both methods perform when confronted with 1) gene sequence or 2) subtype class not present in the training set.
The first validation was a leave-one-out cross-validation test that iterated through each gene in the training set, retraining the prediction tools on a training set which excludes the selected test gene, and then confirming if the retrained predictor could recover the subtype of the test gene.
This validation tests how the predictors perform when run on a distinct sequence that is not in the training set.
The second validation examined how the predictors respond when presented with a gene that has a subtype not found in the reference set.
In this validation, each subtype was iterated over and all genes that are assigned the subtype were removed from the training set.
In each iteration, we recorded the number of false-positive subtype assignments when the test sequences were used as input.
The correct response for the predictors was to return a negative result, since the subtype does not exist in the training set.
For these assessments, we developed a sequence-similarity based tool that assigns putative subtypes using BLAST.
Details for the BLAST tool and how the assessment was conducted are available in supplementary methods section.
The assessment examined the five subtype schemes available in Phylotyper: Stx1, Stx2, Eae, H-type (FliC), O-type (Wzy \& Wzx).
When tasked with assigning a novel gene sequence (i.e. one that is not in the training set) in the leave-one-out validation, Phylotyper consistently had higher precision than a top-BLAST hit approach.
The average precision in Phylotyper was 0.99 versus 0.96 in the top-BLAST hit approach(result details are available in Supplementary Table S2).
In order to achieve this level of precision, the BLAST approach often had lower recall rates; it had an average recall of 0.81 compared to 0.90 with Phylotyper
Similarily, when entire subtype classes were withheld from the training set, Phylotyper had consistenly lower false positive rates for all subtypes schemes tested; the average precision in this challenging test case was 0.11.
In the BLAST approach, the average precision was 0.30.

A separate assessment for the V-typer tool; a Stx subtype predictor, was run using selected Stx gene sequences from the experimentally-verified Phylotyper training set.
The test Stx genes had sufficent surrounding DNA sequence to support \it{in silico} PCR.
In total, 24 Stx gene sequences were tested with the V-typer tool and V-typer returned results for 7, all correct.
Phylotyper can correctly predict the subtype for all these genes.
Based on this level of recall, it appears the conditions in the Stx subtype setting are challenging for simulated PCR.

@@

Early developers of a serotype predictor for E. coli. This tool uses BLAST.
Performance testing showed that Phylotyper is more robust than an approach based solely on sequence similarity.
The goal of this tool is allele matching, and dependence on reference training set is critical to performance.
To implement allele matching is relatively straight-forward, although some subtyping tools operate with input data such as short reads that require careful analysis.
EcOH uses SRST tool to align and match short reads from ... to known annotated sequences. Again the focus is on matching rather than inference.

Phylotyper computes an empirical model of subtype evolution to predict subtypes for unclassified sequences.
By estimating the phylogenetic distribution of each subtype, Phylotyper is less likely to make type 1 error when encountering a novel subtype. Our empircal testing confirmed this behavior. 
The rate of false positive classifications when we withheld an entire subtype from the training set and used as it the test set was significantly lower than with a straight-forward

### 2017-05-10 manuscript/phylotyper_suppl_info.tex
Updated run-time for O-types (original test genome had 4 copies of O-antigen cluster resulting in slower run-time)
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@

\begin{minipage}{\linewidth}
\setlength{\tabcolsep}{4pt}
\captionof{table}{Leave-One-Out Cross Validation Results}
\medskip
\begin{tabular}{@{\extracolsep{4pt}}llll>{\centering}m{1.6cm}lll@{}}
\toprule 
\multirow{2}{*}{Scheme} & \multicolumn{4}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
\cline{2-5}\cline{6-8}
& Precision & Recall & F$_{1}$ Score & Run-time (s) $^{1}$ & Precision & Recall & F$_{1}$ Score \\
\midrule
{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 6 & 0.94 & 0.94 & 0.94\\
{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 32 & 0.93 & 0.93 & 0.93\\
{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 17 & 0.99 & 0.98 & 0.99\\
{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 278 & 1.00 & 0.36 & 0.53\\\bottomrule
\end{tabular}\par
\bigskip
Formula:
\begin{enumerate}
\item $Precision = TP / (TP + FP)$
\item $Recall = TP / (TP + FN)$
\item $F_{1}~score = 2*Precision*Recall/(Precision + Recall)$
\end{enumerate}
$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
\par
\bigskip
$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a non-identical subtype gene which triggers the full phylotyper pipeline.

>>>>

\section{Supplementary Methods}

\subsection{Performance Analysis}

\subsubsection{BLAST-based method}

BLAST

\subsubsection{Selecting Percent Identity Cutoff for BLAST Approach}

Following \citep{Jenkins2015}, percent identity in the BLAST alignments is used as a cutoff. Subtype classifications are transfered from the top BLAST match provided the alignment percent identity is above the cutoff. The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g. type I error), too high and valid subtype allele BLAST hits are filtered out (e.g. type II error). To select the ideal cutoff that balances these competing sources of error, a cross-validation simulation was run to estimate false discovery and true positive rates across varying percent identity cutoffs. A 10-fold cross-validation was performed; 10\% of the training set was randomly selected. The remaining 90\% was used to build the reference BLAST database. The 10\% test set was then searched against this database using BLAST and positive and negative results were recorded. This step was repeated 100 times and the F1-score, which is a weighted average of the precision and recall, was calculated. Precision is defined as:

$Precision = TP / (TP + FP)$

and recall as:

$Recall = TP / (TP + FN)$

These values are then combined to produce an F-score as follows:

$2 * ( precision * recall ) / ( precision + recall )$

The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.  F-score was computed for increasing percent identities, and the percent identity corresponding to the maximum F-score was selected as the cutoff.

\subsubsection{Performance Assessment}

Two separate analyses

@@

%\captionsetup{justification=RaggedRight}

\begin{minipage}{\linewidth}
\centering
\setlength{\tabcolsep}{4pt}
\captionof{table}[justification=RaggedRight]{Leave-One-Out Cross Validation Results}
\medskip
\begin{tabular}{@{\extracolsep{4pt}}llll>{\centering}m{1.6cm}lll@{}}
\toprule 
\multirow{2}{*}{Scheme} & \multicolumn{4}{c}{Phylotyper} & \multicolumn{3}{c}{Sequence-similarity}\\
\cline{2-5}\cline{6-8}
& Precision & Recall & F$_{1}$ Score & Run-time (s) $^{1}$ & Precision & Recall & F$_{1}$ Score \\
\midrule
{\it E. coli} Stx1 & 1.00 & 0.94 & 0.97 & 6 & 0.94 & 0.94 & 0.94\\
{\it E. coli} Stx2 & 1.00 & 0.99 & 0.99 & 32 & 0.93 & 0.93 & 0.93\\
{\it E. coli} Intimin & 1.00 & 0.98 & 0.99 & 17 & 0.99 & 0.98 & 0.99\\
{\it E. coli} H-serotype & 0.99 & 0.98 & 0.98 & 16 & 0.96 & 0.85 & 0.90\\
{\it E. coli} O-serotype & 1.00 & 0.61 & 0.75 & 67 & 1.00 & 0.36 & 0.53\\\bottomrule
\end{tabular}\par
\bigskip
\raggedright
Formula:
\begin{enumerate}
\item $Precision = TP / (TP + FP)$
\item $Recall = TP / (TP + FN)$
\item $F_{1}~score = 2*Precision*Recall/(Precision + Recall)$
\end{enumerate}
$TP = True~Positive,~~FP = False~Positive,~~FN = False~Negative$
\par
\bigskip
$^{1}$ Run-time is for an {\it Escherichia coli} genome containing a non-identical subtype gene which triggers the full phylotyper pipeline.

### 2017-05-10 manuscript/phylotyper_suppl_info_revised.tex
UPDATE: new figure colors, ADD: supplementary table S3, ADD: Results section

@@

\begin{minipage}{\linewidth}
\centering
\captionof{table}{V-typer Performance}
\medskip
\begin{tabular}{@{}lll@{}}\toprule
NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ \\\midrule
CP015020.1 & 2d & 2d & N \\
stx2 & Shiga-toxin 1 subtype & {\it E. coli} & 2\\
eae & Intimin subtype & {\it E. coli} & 1\\
flic & H-serotype based on flagellin gene; fliC & {\it E. coli} & 1\\
wz & O-serotype based on the wzy and wzx genes & {\it E. coli} & 2\\\bottomrule
\end{tabular}\par
\bigskip
\raggedright
$^{1}$ N=Negative (no results returned)

>>>>

\section{Supplementary Methods}

\subsection{Performance Analysis}

\subsubsection{BLAST-based method}

BLAST

\subsubsection{Selecting Percent Identity Cutoff for BLAST Approach}

Following \citep{Jenkins2015}, percent identity in the BLAST alignments is used as a cutoff. Subtype classifications are transfered from the top BLAST match provided the alignment percent identity is above the cutoff. The cutoff impacts performance in one of two ways: too low and non-equivalent subtype alleles may be matched with the query (e.g. type I error), too high and valid subtype allele BLAST hits are filtered out (e.g. type II error). To select the ideal cutoff that balances these competing sources of error, a cross-validation simulation was run to estimate false discovery and true positive rates across varying percent identity cutoffs. A 10-fold cross-validation was performed; 10\% of the training set was randomly selected. The remaining 90\% was used to build the reference BLAST database. The 10\% test set was then searched against this database using BLAST and positive and negative results were recorded. This step was repeated 100 times and the F1-score, which is a weighted average of the precision and recall, was calculated. Precision is defined as:

$Precision = TP / (TP + FP)$

and recall as:

$Recall = TP / (TP + FN)$

These values are then combined to produce an F-score as follows:

$2 * ( precision * recall ) / ( precision + recall )$

The F-score; a binary classifer metric, was recorded individually for each subtype and then averaged to create a single performance metric for this multi-class predictor.  F-score was computed for increasing percent identities, and the percent identity corresponding to the maximum F-score was selected as the cutoff.

\subsubsection{Performance Assessment}

Two separate cross-validation tests were performed to compare the performance of Phylotyper with the BLAST method. The comparison to vtyper is a separate analysis, as vtyper does not use a training set and a cross-validation test would not work in this situation. A leave-one-out cross-validation analysis was performed where each sequence in the training set is withheld and the remaining sequences are used for form the phylogenetic tree for Phylotyper or are used to build the BLAST database for the BLAST-based approach. The predicted result for the test sequence is then compared to the true value.

\subsubsection{V-typer Assessment}

The V-typer tool predicts Stx1 or Stx2 subtypes by simulating PCR (i.e. mapping primers). Due to the unique method in V-typer,  which does not use a training dataset, the cross-validation assessment was adapted to test V-typer. Each experimentally-verified Stx gene sequence that makes up Phylotyper training dataset was searched for against the NCBI nt database to retrieve surrounding sequence that may be involved in the PCR detection by V-typer.  Specifically, regions that have at least 5 kB up- and downstream from the Stx gene were selected for the assessment. V-typer was run on each sequence and the results recorded in supplmentary table XXXX.

@@

\begin{minipage}{\linewidth}
\centering
\captionof{table}{V-typer Performance}
\medskip
\begin{tabular}{@{}llll@{}}\toprule
NCBI Accession & Stx Subtype & V-typer Prediciton$^{1}$ & Phylotyper Prediction\\\midrule
CP015020 & 2d & N & 2d \\
CP015229 & 2g & 2g & 2g \\
CP009106 & 2a & N & 2a \\
CP007133 & 2a & 2a & 2a \\
CP006262 & 2a & 2a & 2a \\
AP010960 & 2a & 2a & 2a \\
CP013663 & 2a & N & 2a \\
CP015228 & 2a & N & 2a \\
CP011331 & 2a & N & 2a \\
HF572917 & 2a & N & 2a \\
LN554923 & 2a & N & 2a \\
LM997071 & 2a & N & 2a \\
CP018250 & 2c & N & 2c \\
CP018252 & 2c & N & 2c \\
CP018247 & 2c & N & 2c \\
CP018245 & 2c & N & 2c \\
CP018243 & 2c & N & 2c \\
CP017446 & 2c & N & 2c \\
CP017442 & 2c & N & 2c \\
CP017438 & 2c & N & 2c \\
CP014314 & 2c & N & 2c \\
LM997036 & 2a & 2a & 2a \\
LM996489 & 1c & 1c & 1c \\
LM997036 & 1c & 1c & 1c \\
Total & 24 & 7 & 24 \\
\bottomrule
\end{tabular}\par
\bigskip
\raggedright
$^{1}$ N=Negative (no results returned)
